# Reproducible Research: Peer Assessment 1
January 2015, Andreas von Ballmoos

This is an R markdown document to solve the given assignment.

## Prerequisites
First I make sure that the knitr and ggplot2 packages are installed and loaded. I take the opportunity to get used to plot with ggplot2.

```{r}
if (!require("knitr")) {
install.packages("knitr")
}
require("knitr") 

if (!require("ggplot2")) {
install.packages("ggplot2")
}
require("ggplot2") 
````

## Loading and preprocessing the data
Before loading the data, I unzip the source file. I then read it in and convert the data column to date format.
```{r}
unzip("activity.zip")
data <- read.csv("activity.csv", na.strings="NA")
data$date=as.Date(data$date)
```


## What is mean total number of steps taken per day?
### Make a histogram of the total number of steps taken each day
```{r}
qplot(data$steps, binwidth=50)
```

### Calculate and report the mean and median total number of steps taken per day
Before the calculating, get rid of the NAs
```{r}
good <- !is.na(data$steps)
data2 <- data[good,]

total_steps_by_day <- aggregate(data2$steps, list(data2$date), sum)
mean(total_steps_by_day$x)
median(total_steps_by_day$x)
```

## What is the average daily activity pattern?
### Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days
```{r}
good <- !is.na(data$steps)
data2 <- data[good,]

mean_steps_by_interval <- aggregate(data2$steps, list(data2$interval), mean)
g <- ggplot(mean_steps_by_interval, aes(x=mean_steps_by_interval$Group.1, y=mean_steps_by_interval$x))
g + geom_line()
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
Looking at the plot above, the interval must be around 800
```{r}
y <- mean_steps_by_interval[order(mean_steps_by_interval$x, na.last=T, decreasing =T),]
head(y,1)
```
It's the interval with the id 835

## Imputing missing values
### Calculate and report the total number of missing values in the dataset
There are 2304 NA's based on the summary() function
```{r}
summary(data)
summary(good)
```
### Devise a strategy for filling in all of the missing values in the dataset
I think the mean of the interval rounded to an integer is a good value to replace the NAs with. The values of a given interval are quite stable over the days.

### Create a new dataset that is equal to the original dataset but with the missing data filled in
First merge the dataset with the mean of the steps by interval - which serves as default value for an interval - and sort it back.
```{r}
data3 <- merge(data, mean_steps_by_interval, by.x="interval", by.y="Group.1")
data4 <- data3[with(data3, order(date, interval)), ]
```
Now loop through the dataset and replace NAs with the default values
```{r}
for (i in 1:nrow(data4) ) {
    if (is.na(data4[i,2])) data4[i,2] <- data4[i,4]
}
head(data4)
```
And remove the column with the default values and put the columns in the original order again:
```{r}
steps <- as.integer(round(data4[,2], digits=0))
date <- data4[,3]
interval <- data4[,1]
data5 <- data.frame(steps, date, interval)
head(data5)
```
### Make a histogram of the total number of steps taken each day
```{r}
qplot(data5$steps, binwidth=50)
```

### Calculate and report the mean and median total number of steps taken per day
```{r}
total_steps_by_day <- aggregate(data5$steps, list(data5$date), sum)
mean(total_steps_by_day$x)
median(total_steps_by_day$x)
```
The numbers differ a little bit from the ones that were calculated without the NA's. But the difference is very small (< 0.1%).

### What is the impact of imputing missing data on the estimates of the total daily number of steps?

*Keine Ahnung was das soll*

## Are there differences in activity patterns between weekdays and weekends?
To answer this questions we need to add grouping information to the dataset that indicates whether a day is a weekday or a weekend day.
### Create a new factor variable in the dataset with two levels
```{r}
Sys.setlocale("LC_TIME", "English")
data5$dayofweek <- weekdays(data5$date, abbreviate=T)
head(data5)
```
And loop again through the dataset to determine weekend days:
```{r}
data5$weekdaycategory <- "weekday"
for (i in 1:nrow(data5) ) {
    if (data5[i,4]=='Sat') data5[i,5] <- 'weekend'
    if (data5[i,4]=='Sun') data5[i,5] <- 'weekend'
}
data5$weekdaycategory <- factor(data5$weekdaycategory)
str(data5)
```

### Make a panel plot containing a time series plot
Split the graph into weekdays/weekends with the average steps taken per interval:
```{r}
g2 <- ggplot(data=data5, mapping=aes(x=interval, y=steps))
g2 + facet_grid(weekdaycategory ~ .) + layer(geom = "line",
                  stat = "summary",
                  fun.y = "mean")
```

Uff... This was an extremly hard piece!
